Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings

### 2021-06-29 reviews/response.md

In the following response, I would like to address each of your comments
and highlight the amendments we have made to the manuscript and the liminal software.

# Reviewer A

> Unfortunately, I was not able to run the provided code. My guess is that there is some incompatibility with the newest R-version 4-0-3. Please check!

This was a problem with a newer version of the tourr package causing liminal
to not animate the tour correctly. It has been fixed and the liminal package is now 
available on CRAN and has been tested against several versions of R on multiple
operating systems.

## Major remarks

> The manuscript is well written and presents the approach in a balanced way. Despite
this, the goal and primary focus of the paper gets not always clear as it it switches focus
between presenting an enhancement of NLDR methods, presenting a better alternative to
NLDR methods for certain tasks (clustering) and describing the general functionality of
interactions with grand tour representations. A revision should clarify the goals of the
manuscript.
> Secondly, the relevance of having a cluster orientation task at hand seems relevant for
some arguments but not for others. Here also, the authors should clearly provide guidance
to the reader on which data and research contexts are essential for certain conclusions.

I have added comments to the abstract and throughout the manuscript
to clarify that the goal of the liminal interface is to both correct an embedding
(as shown in case study 2) and for clustering (case study 3). The discussion
also clarifies the goals of the method.

> Thirdly, the term high-dimensional data has gained fairly specfic meaning in some
contexts. Please clarify whether you use the term in its rather broad sense or whether you
want to allude particularly to small-N-large-p situations and their resulting conditions.

This has been clarified in the introduction.

> Fourthly, the manuscript has been enhanced by links to video sequences of tour ani-
mations for the case studies. While these videos add a lot to the static description in the
papers, additional explanations to the videos (voice-over or captions) would further ease
the understanding of the animations and hence improve further the manuscript.

Thank you for the suggestion, I have now added descriptions and
captions and descriptions to all videos.

> Minor remarks

These have all been amended in the text.

@@

# Reviewer B

> For example, the various clusters came pre-colored.  I would like to see a little more about how the analyst would identify these clusters, especially in a case such as your tree example where it doesn't appear that you would have the coloring/clustering you have without a priori knowledge of the different branches.  Is it possible to create that clustering starting from the tour or the embedding?

This is possible, but we considered it out of scope for the paper to go into
specifics of how one would perform a cluster analysis. 
I've added some sentences to the main text about how one might go about using
the interface to generate clusters manually..  The GGobi book by the last author of the manuscript details how one might spin and brush using a tour, to 
build up a cluster analysis manually.

@@

> Would it be worthwhile to link a pairwise scatterplot matrix for all or a subset of the original variables or components so as to be able to relate clusters and other features via brushing back to those variables for interpretation?  I guess the biplot in Figure 7 may be intended to show something similar?  Is it possible to expand and clarify your results from this figure a bit for those of us who are not genomics specialists?

I have expanded what specific terms mean in the main text, to clarify the purposes
of this type of analysis and the graphic. Essentially marker genes are of interest
to biologists for determining signatures that represent cells. The biplot
on this view was intended to show how the relative expression of each gene could
differentiate between possible cell types (clusters). We didn't use a
scatterplot matrix here because it would be difficult to read with the large number
of variables. As the liminal interface is a shiny app, it could be hooked up to
other visualisations by the user.

Yours Sincerely,

Dr Stuart Lee

### 2021-06-29 liminal.Rmd
update main text and email

---
documentclass: jdssv
author:
  - name: Stuart Lee
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    email: \email{stuart.lee1@monash.edu}
  - name: Ursula Laa 
    affiliation: Monash University
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    # To add another line, use \AND at the end of the previous one as above
  - name: Dianne Cook
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    affiliation: Monash University
title:
  formatted: "Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings"
  # If you use tex in the formatted title, also supply version without
  plain:     "Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings"
  # For running headers, if needed
  short:     "\\pkg{liminal}: High-Dimensional Interactive Data Visualisation"
abstract: >
  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complex parameterisations; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique. We present visual diagnostics for the pragmatic usage of NLDR methods by combining them with a technique called the tour. A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space. The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view. By linking the tour to an NLDR view, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading. We display several case studies from both simulations and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks and for correcting an NLDR method. The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available on CRAN.
keywords:
  # at least one keyword must be supplied
  formatted: [dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, "\\proglang{R}"]
  plain:     [dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, R]
output: 
  bookdown::pdf_document2:
    template: template/jdssv_template.tex
    keep_tex: true
    toc: false
    citation_package: natbib
bibliography: liminal.bib
biblio-style: apalike
---
```{r setup, include = FALSE}
options(tinytex.clean = FALSE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.show = 'hold',
  dpi = 300,
  out.width = "\\textwidth",
  fig.path = here::here('img/')
)
```

@@

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed quality control using the **scater** package
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using **scran** 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters [@Blondel2008-bx].

@@

Due to latency of the \pkg{liminal} interface we do a weighted sample of  the rows based on
cluster membership, leaving us with approximately 10 per cent of the original
data size - 4,590 cells. Although this is not ideal, it still allows us to
get a sense of the shape of the clusters as seen from the linked video in Figure \@ref(fig:mouse-01). If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the \pkg{liminal} view as a way finding more granular cluster structure.

@@

Next, we identify marker genes for clusters using one sided Welch t-tests with
a minimum log fold change of one as recommended by @Amezquita2020-at, which
uses the testing framework from @McCarthy2009-qx. We select the top 10 marker 
genes that are upregulated in cluster 2, which was one of the clumped clusters
when we toured on principal components. Here the tour becomes an alternative
to a standard heatmap view for assessing shared markers; the basis
generation (shown as the biplot on the left view) reflects the relative 
weighting of each gene. We run the tour directly on the 
log normalised expression values using the same subset as before.

>>>>

---
documentclass: jdssv
author:
  - name: Stuart Lee
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    email: \email{stuart.andrew.lee@gmail.com}
  - name: Ursula Laa 
    affiliation: Monash University
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    # To add another line, use \AND at the end of the previous one as above
  - name: Dianne Cook
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    affiliation: Monash University
title:
  formatted: "Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings"
  # If you use tex in the formatted title, also supply version without
  plain:     "Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings"
  # For running headers, if needed
  short:     "\\pkg{liminal}: High-Dimensional Interactive Data Visualisation"
abstract: >
  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complexity; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique. We present visual diagnostics for the usage of NLDR methods by combining them with a technique called the tour. A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space. The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view. By linking the tour to an NLDR view, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading. We display several case studies from both simulations and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks and for correcting an NLDR embedding. The implementation of our framework is available as an \proglang{R} package called \pkg{liminal}.
keywords:
  # at least one keyword must be supplied
  formatted: [dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, "\\proglang{R}"]
  plain:     [dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, R]
output: 
  bookdown::pdf_document2:
    template: template/jdssv_template.tex
    keep_tex: true
    toc: false
    citation_package: natbib
bibliography: liminal.bib
biblio-style: apalike
---
```{r setup, include = FALSE}
options(tinytex.clean = FALSE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.show = 'hold',
  dpi = 300,
  out.width = "\\textwidth",
  fig.path = here::here('img/')
)
```

@@

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed quality control using the **scater** package
by removing cells with high proportion of mitochondrial gene expression (as this indicates poor sample quality) and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using **scran** 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters [@Blondel2008-bx].

@@

Due to latency of the \pkg{liminal} interface we do a weighted sample of  the rows based on
cluster membership, leaving us with approximately 10 per cent of the original
data size - 4,590 cells. Although this is not ideal, it still allows us to
get a sense of the shape of the clusters as seen from the linked video in Figure \@ref(fig:mouse-01). If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the \pkg{liminal} view as a way finding more granular cluster structure. One could perform this approach manually via the \pkg{liminal} interface by returning the regions identified by brushing on the tour or embedding view.

@@

Next, we identify marker genes for clusters using one sided Welch t-tests with
a minimum log fold change of one as recommended by @Amezquita2020-at, which
uses the testing framework from @McCarthy2009-qx. We select the top 10 marker 
genes that are expressed in relatively larger quantities (upregulated) in cluster 2, which was one of the clumped clusters
when we toured on principal components. Here the tour becomes an alternative
to a standard heatmap view for assessing shared markers; the basis
generation (shown as the biplot on the left view) reflects the relative 
weighting of each gene. We run the tour directly on the 
log normalised expression values using the same subset as before.
